Transcriptomic responses to diet quality and viral infection in Apis mellifera

### 2018-09-24 Manuscript/Text/manuscript.tex
Bonning corrections

\title{Gene expression responses to diet quality and viral infection in Apis mellifera}

@@

\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
   email={lrutter@iastate.edu}   % email address
]{\inits{LR}\fnm{Lindsay} \snm{Rutter}}
\author[
   addressref={aff2},
   email={dicook@monash.edu}
]{\inits{DC}\fnm{Dianne} \snm{Cook}}
\author[
   addressref={aff3,aff4},
   email={amytoth@iastate.edu}
]{\inits{AT}\fnm{Amy L.} \snm{Toth}}
\author[
   addressref={aff5},
   corref={aff5},
   email={adolezal@illinois.edu}
]{\inits{AD}\fnm{Adam} \snm{Dolezal}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors' addresses here        %%
%%                                          %%
%% Repeat \address commands as much as      %%
%% required.                                %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

@@

\begin{abstract}
\parttitle{Background}
Parts of Europe and the United States have witnessed dramatic losses in commercially managed honey bees over the past decade to what is considered an unsustainable extent. The large-scale loss of honey bees has considerable implications for the agricultural economy because honey bees are one of the leading pollinators of numerous crops. Honey bee declines have been associated with several interactive factors. Poor nutrition and viral infection are two environmental stressors that pose heightened dangers to honey bee health. In this study, we used RNA-sequencing to examine how monofloral diets and Israeli Acute Paralysis Virus inoculation influence gene expression patterns in honey bees.

@@

Commercially managed honey bees have undergone unusually large declines in the United States and parts of Europe over the past decade \cite{ccd1, ccd2, ccd3}, with annual mortality rates exceeding what beekeepers consider sustainable \cite{ccd5, ccd6}. More than 70 percent of major global food crops (including fruits, vegetables, and nuts) at least benefit from pollination, and yearly insect pollination services are valued worldwide at \$175 billion \cite{ccd7}. As honey bees are largely considered to be the leading pollinator of numerous crops, their marked loss has considerable implications regarding agricultural sustainability \cite{ccd4}.

Honey bee declines have been associated with several factors, including pesticide use, parasites, pathogens, habitat loss, and poor nutrition \cite{factors, factors2}. Researchers generally agree that these stressors do not act in isolation; instead, they appear to influence the large-scale loss of honey bees in interactive fashions as the environment changes \cite{interacting}. Nutrition and viral infection are two broad factors that pose heightened dangers to honey bee health in response to recent environmental changes.

Pollen is a main source of nutrition (including proteins, amino acids, lipids, sterols, starch, vitamins, and minerals) in honey bees \cite{source, source2}. At the individual level, pollen supplies most of the nutrients necessary for physiological development \cite{brodschneider} and is believed to have considerable impact on longevity \cite{longevity}. At the colony level, pollen enables young workers to produce jelly, which then nourishes larvae, drones, older workers, and the queen \cite{jelly, jelly2}. Various environmental changes (including urbanization and monoculture crop production) have significantly altered the nutritional profile available to honey bees. In particular, honey bees are confronted with less diverse selections of pollen, which is of concern because mixed-pollen (polyfloral) diets are generally considered healthier than single-pollen (monofloral) diets \cite{diverse, diverse2, alaux}. Indeed, reported colony mortality rates are higher in developed land areas compared to undeveloped land areas \cite{undeveloped}, and beekeepers rank poor nutrition as one of the main reasons for colony losses \cite{bkLoss}. Understanding how undiversified diets affect honey bee health will be crucial to resolve problems that may arise as agriculture continues to intensify throughout the world \cite{ag, ag2}.

Viral infection was a comparatively minor problem in honey bees until the last century when Varroa destructor (an ectoparasitic mite) spread worldwide \cite{miteSpread}. This mite feeds on honey bee hemolymph \cite{hemolymph}, transmits multiple viruses, and supports replication of certain viruses \cite{miteVirus, miteVirus2, miteVirus3}. More than 20 honey bee viruses have been identified \cite{numVirus}. One of these viruses that has been linked to honey bee decline is Israeli Acute Paralysis Virus (IAPV). A positive-sense RNA virus of the Dicistroviridae family \cite{fam}, IAPV causes infected honey bees to display shivering wings, decreased locomotion, muscle spams, paralysis, and high premature death percentages in caged infected adult honey bees \cite{symptoms}. IAPV has demonstrated higher infectious capacities than other honey bee viruses in certain conditions \cite{carrillo} and is more prevalent in colonies that do not survive the winter \cite{winter}. Its role in the rising phenomenon of ``Colony Collapse Disorder'' (in which the majority of worker bees disappear from a hive) remains unclear: It has been implicated in some studies \cite{iapvCCD, iapvCCD2} but not in other studies \cite{ccd1, fam, iapvCCD3}. Nonetheless, it is clear that IAPV reduces colony strength and survival.

Although there is growing interest in how viruses and diet quality affect the health and sustainability of honey bees, as well as a recognition that such factors might operate interactively, there are only a small number of experimental studies thus far directed toward elucidating the interactive effects of these two factors in honey bees \cite{intNV, intNV2, intNV3}. We recently used laboratory cages and nucleus hive experiments to investigate the health effects of these two factors, and our results show the importance of the combined effects of both diet quality and virus infection. Specifically, high quality pollen is able to mitigate virus-induced mortality to the level of diverse, polyfloral pollen \cite{adamInt}.

Following up on these phenotypic findings from our previous study, we now aim to understand the corresponding underlying mechanisms by which high quality diets protect bees from virus-induced mortality. For example, it is not known whether the protective effect of good diet is due to direct, specific effects on immune function (resistance), or if it is due to indirect effects of good nutrition on vigor (tolerance) \cite{resTol1}. Transcriptomics is one means to better understand the mechanistic underpinnings of dietary and viral effects on honey bee health. Transcriptomic analysis can help us identify 1) the genomic scale of transcriptomic response to diet and virus infection, 2) whether these factors interact in an additive or synergistic way on transcriptome function, and 3) the types of pathways affected by diet quality and viral infection. This information, heretofore lacking in the literature, can help us better understand how good nutrition may be able to serve as a ``buffer'' against other stressors \cite{AdamTothReview}. As it stands, there are only a small number of published experiments examining gene expression patterns related to diet effects \cite{alaux2} and virus infection effects \cite{galbraith} in honey bees. As far as we know, there are few to no studies investigating honey bee gene expression patterns specifically related to monofloral diets, and few to no studies investigating honey bee gene expression patterns related to the combined effects of diet in any broad sense and viral inoculation in any broad sense.

In this study, we examine how monofloral diets and viral inoculation influence gene expression patterns in honey bees by focusing on four treatment groups (low quality diet without IAPV exposure, high quality diet without IAPV exposure, low quality diet with IAPV exposure, and high quality diet with IAPV exposure). We conduct RNA-sequencing analysis on a randomly selected subset of the honey bees we used in our previous study (as is further described in our methods section). We then examine pairwise combinations of treatment groups, the main effect of monofloral diet, the main effect of IAPV exposure, and the combined effect of the two factors on gene expression patterns.

@@

As shown in Figure 2, IAPV titers of honey bees 72 hour post-inoculation significantly differed among the treatment groups (mixed model ANOVA across all treatment groups, df = 3, 33; F = 6.10; p $<$ 2.03e-03). The effect of virus treatment (mixed model ANOVA, df = 1, 33; F = 15.04; p $<$ 4.75e-04) was significant, but the diet treatment (mixed model ANOVA, df = 1, 33; F = 2.55; p = 1.20e-01) and the interaction between the two factors (mixed model ANOVA, df = 1, 33; F = 7.02e-01, p = 4.08e-01) were not significant. We compared IAPV titer volumes  based on pairwise comparisons: Bees fed Rockrose pollen had significantly elevated IAPV titer volumes with virus infection compared to uninfected controls (Benjamini Hochberg, p $<$ 7.56e-03). However, bees fed Chestnut pollen did not have significantly elevated IAPV titer volumes with virus infection compared to uninfected controls (Benjamini Hochberg, p = 6.29e-02). Overall, we interpreted these findings to mean that high-quality Chestnut pollen could ``rescue'' high virus titers resulting from the inoculation treatment, whereas low-quality Rockrose pollen could not do so (Figure 2).

@@

We observed a substantially larger number of differentially expressed genes (DEGs) in our diet main effect (\textit{n} = 1,914) than in our virus main effect (\textit{n} = 43) (Supplementary table 1 A and B, Additional file 1). In the diet factor, there were more Chestnut-upregulated DEGs (\textit{n} = 1033) than Rockrose-upregulated DEGs (\textit{n} = 881). In the virus factor, there were more virus-upregulated DEGs (\textit{n} = 38) than control-upregulated DEGs (\textit{n} = 5). While these reported DEG counts are from the DESeq2 package, we saw similar trends for the edgeR and limma package results (Supplementary table 1, Additional file 1 and Additional file 18).

@@

We wished to explore the signal:to:noise ratio between the Galbraith dataset and our dataset. Basic PCA plots were constructed with the DESeq2 analysis pipeline and showed that the Galbraith dataset may better separate the infected and uninfected honey bees better than our dataset (Additional file 2). We also noted that the first replicate of both treatment groups in the Galbraith data did not cluster as cleanly in the PCA plots. However, through this automatically-generated plot, we can only visualize information at the sample level. Wanting to learn more about the data at the gene level, we continued with additional visualization techniques.

We used parallel coordinate lines superimposed onto boxplots to visualize the DEGs associated with virus infection in the two studies. The background boxplot represents the distribution of all genes in the data, and each parallel coordinate line represents one DEG. To reduce overplotting of parallel coordinate lines, we used hierarchical clustering techniques to separate DEGs into common patterns as is described in the methods section.

@@

Finally, we looked at scatterplot matrices to assess the DEGs. We created standardized scatterplot matrices for each of the four clusters (Figure 3) of the Galbraith data (Additional files 7, 8, 9, and 10). We also created standardized scatterplot matrices for our data. However, as our dataset contained 24 samples, we would need to include 276 scatterplots in our matrix, which would be too numerous to allow for efficient visual assessment of the data. As a result, we created four scatterplot matrices of our data, each with subsets of 6 samples to be more comparable to the Galbraith data (Additional files 11, 12, 13, and 14). We can again confirm through these plots that the DEGs from the Galbraith data appeared more as expected: Deviating more from the \textit{x=y} line in the treatment scatterplots while staying close to the \textit{x=y} line in replicate scatterplots.

@@

Using the contrasts specified in Table \ref{tbl:contrasts}, we discovered 122 ``tolerance'' candidate DEGs and 125 ``resistance'' candidate DEGs. We again used parallel coordinate lines superimposed onto boxplots to visualize these candidate DEGs. To reduce overplotting of parallel coordinate lines, we again used hierarchical clustering techniques to separate DEGs into common patterns. Perhaps unsurprisingly, we still see a substantial amount of noise (inconsistency between replicates) in our resulting candidate DEGs (Additional files 15 and 16). However, the broad patterns we expect to see still emerge: For example, based on the contrasts we created to obtain the `tolerance'' candidate DEGs, we expect them to display larger count values in the ``NC'' group compared to the ``NR'' group and larger count values in the ``VC'' group compared to the ``VR'' group. Indeed, we see this pattern in the associated parallel coordinate plots (Additional file 15). Likewise, based on the contrasts we created to obtain the `resistance'' candidate DEGs, we still expect them to display larger count values in the ``VC'' group compared to the ``VR'' group, but we no longer expect to see a difference between the ``NC'' and ``NR'' groups. We do generally see these expected patterns in the associated parallel coordinate plots: While there are large outliers in the ``NC'' group, the ``NR'' replicates are no longer typically below a standardized count of zero (Additional file 16). The genes in Cluster 3 may follow the expected pattern the most distinctively (Additional file 16).

@@

In general, the R-squared values between gene read counts and pathogen response measurements were low (R-squared $<$ 0.1). However, some DEG clusters showed slightly larger R-squared values than the non-DEG group (the rest of the data). One prominent example of this includes the first and second cluster of the virus-related DEGs (Additional file 19I). The Kruskal–Wallis test was used to determine if R-squared populations of DEG clusters significantly differed from those of the rest of the data. The p-values and Bonferroni correction values for each of the 36 tests (as described in the methods section) is provided in Supplementary table 9, Additional file 1. An overall trend emerges to suggest that DEGs may have significantly larger correlation with the pathogen response measurements compared to non-DEGs. It is difficult to interpret these results in light of the noisiness of this data, but it may be of interest to conduct further studies examining differential expression between pathogen response measurements.

@@

While many other insect systems show relatively low transcriptional responses to virus infection, previous work on honey bees has revealed hundreds of DEGs \cite{galbraith}. However, the transcriptomic response to virus infection in our experiment was fairly limited. We found only 43 transcripts to be differentially expressed, some with known immune functions (Table \ref{tbl:virusGenes}) such as argonaute-2 and a gene with similarity to MD-2 lipid recognition protein, as well as genes related to transcriptional regulation and muscle contraction. The small number of DEGs in this study may be partly explained by the large amount of noise in the data (Figure 4 and Additional files 2B, 4, 11, 12, 13, and 14).

@@

One of the goals of this study was to use our RNA-sequencing data to assess whether transcriptomic responses to diet quality and virus infection provide insight into whether high quality diet can buffer bees from pathogen stress via mechanisms of ``resistance'' or ``tolerance''. Recent evidence has suggested that overall immunity is determined by more than just ``resistance'' (the reduction of pathogen fitness within the host by mechanisms of avoidance and control) \cite{resTol2}. Instead, overall immunity is related to ``resistance'' in conjunction with ``tolerance'' (the reduction of adverse effects and disease resulting from pathogens by mechanisms of healing) \cite{resTol1, resTol2}. Immune-mediated resistance and diet-driven tolerance mechanisms are costly and may compete with each other \cite{resTol1, resTol4}. Data and models have suggested that selection can favor an optimum combination of both resistance and tolerance \cite{resTol5, resTol6, resTol7, resTol8}. We attempted to address this topic through specific gene expression contrasts (Table \ref{tbl:contrasts}), accompanied by GO analysis of the associated gene lists. We found an approximately equal number of resistance (\textit{n} = 125) and tolerance (\textit{n} = 122) related candidate DEGs, suggesting both processes may be playing significant roles in dietary buffering from pathogen induced mortality. Resistance candidate DEGs had functions related to several forms of metabolism (chitin and carbohydrate), regulation of transcription, and cell adhesion. Tolerance candidate DEGs had functions related to carbohydrate metabolism and chitin metabolism; however, they also showed functions related to immune response, including RNA polymerase II transcription and regulation of response to reactive oxygen species (Figure 5A). Previous studies have shown that transcriptional pausing of RNA polymerase II may be an innate immune response in \textit{D. melanogaster} that allows for a more rapid response by increasing the accessibility of promoter regions of virally induced genes \cite{resTol11}. Moreover, circulating haemocytes in insects encapsulate and nodulate pathogens by forming a barrier between the pathogen and the host tissues. This barrier undergoes apoptosis and melanization through the phenoloxidase enzyme cascade, which produces reactive oxygen species \cite{resTol1, resTol9, resTol10}. In insects, melanization has been shown to be toxic to parasites, bacteria, fungi, and lately viruses \cite{melanization}. These possible immunological defense mechanisms within our ``tolerance'' candidate DEGs and metabolic processes within our ``resistance'' candidate DEGs may provide additional evidence of feedbacks between diet and disease in honey bees \cite{AdamTothReview}.

@@

In addition, this study used a whole body RNA-sequencing approach. In future related studies, it may be informative to use tissue-specific methods. Recent evidence has suggested that RNA-sequencing approaches toward composite structures in honey bees leads to false negatives, implying that genes strongly differentially expressed in particular structures may not reach significance within the composite structure. On a similar note, recent studies have also found that within a composite extraction, structures therein may contain opposite patterns of differential expression. We can provide more detailed answers to our original transcriptomic questions if we were to repeat this same experimental design only now at a more refined tissue level \cite{tissueLevel}. Another future direction related to this work would be to integrate multiple omics datasets to investigate monofloral diet quality and IAPV infection in honey bees. Indeed, previous studies in honey bees have found that multiple omics datasets do not always align in a clear-cut manner, and hence may broaden our understanding of the molecular mechanisms being explored \cite{galbraith}.

@@

Details of the procedures we used to prepare virus inoculum, infect and feed caged honey bees, and quantify IAPV can be reviewed in our previous work \cite{adamInt, carrillo}. A linear mixed effects model was used to relate the mortality rates and IAPV titers to the main and interaction effects of the diet and virus factors. The model was fitted to the data by restricted maximum likelihood (REML) using the ``lme'' function in the R package ``nlme''. Post-hoc pairwise comparisons of the four (diet and virus combination) treatment groups were performed and Benjamini-Hochberg adjusted p-values were calculated to limit familywise Type I error rates \cite{bh}.

@@

We used an array of visualization tools as part of our analysis. We first used popular tools like the PCA plot \cite{pca} from the DESeq2 package. After that, we used multivariate visualization tools from our work-in-progress R package called bigPint. Specifically, we used parallel coordinate plots \cite{origPCP}, scatterplot matrices \cite{scatMat}, and litre plots (which we recently developed based on ``replicate line plots'' \cite{jds} (cite bigPint too)) to assess the variability between the replicates and the treatments in our data. We also used these plotting techniques to assess for normalization problems and other common problems in RNA-sequencing analysis pipelines \cite{jds} (cite bigPint too).

>>>>

\title{Transcriptomic responses to diet quality and viral infection in Apis mellifera}

@@

\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
   email={lrutter@iastate.edu}   % email address
]{\inits{LR}\fnm{Lindsay} \snm{Rutter}}
\author[
   addressref={aff6},
   email={bbonning@ufl.edu}
]{\inits{BB}\fnm{Bryony C.} \snm{Bonning}}
\author[
   addressref={aff2},
   email={dicook@monash.edu}
]{\inits{DC}\fnm{Dianne} \snm{Cook}}
\author[
   addressref={aff3,aff4},
   email={amytoth@iastate.edu}
]{\inits{AT}\fnm{Amy L.} \snm{Toth}}
\author[
   addressref={aff5},
   corref={aff5},
   email={adolezal@illinois.edu}
]{\inits{AD}\fnm{Adam} \snm{Dolezal}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors' addresses here        %%
%%                                          %%
%% Repeat \address commands as much as      %%
%% required.                                %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\address[id=aff6]{%
  \orgname{Department of Entomology and Nematology, University of Florida},
  \city{Gainesville},   
  \state{FL}
  \postcode{32611},
  \cny{USA}
}

@@

\begin{abstract}
\parttitle{Background}
Parts of Europe and the United States have witnessed dramatic losses in commercially managed honey bees over the past decade to what is considered an unsustainable extent. The large-scale loss of honey bees has considerable implications for the agricultural economy because honey bees are one of the leading pollinators of numerous crops. Honey bee declines have been associated with several interactive factors. Poor nutrition and viral infection are two environmental stressors that pose heightened dangers to honey bee health. In this study, we used RNA-sequencing to examine how monofloral diets and Israeli acute paralysis virus inoculation influence gene expression patterns in honey bees.

@@

Commercially managed honey bees have undergone unusually large declines in the United States and parts of Europe over the past decade \cite{ccd1, ccd2, ccd3}, with annual mortality rates exceeding what beekeepers consider sustainable \cite{ccd5, ccd6}. More than 70 percent of major global food crops (including fruits, vegetables, and nuts) at least benefit from pollination, and yearly insect pollination services are valued worldwide at \$175 billion \cite{ccd7}. As honey bees are largely considered to be the leading pollinator of numerous crops, their marked loss has considerable implications for agricultural sustainability \cite{ccd4}.

Honey bee declines have been associated with several factors, including pesticide use, parasites, pathogens, habitat loss, and poor nutrition \cite{factors, factors2}. Researchers generally agree that these stressors do not act in isolation; instead, they appear to influence the large-scale loss of honey bees in an interactive fashion as the environment changes \cite{interacting}. Nutrition and viral infection are two broad factors that pose heightened dangers to honey bee health in response to recent environmental changes.

Pollen is a main source of nutrition (including proteins, amino acids, lipids, sterols, starch, vitamins, and minerals) in honey bees \cite{source, source2}. At the individual level, pollen supplies most of the nutrients necessary for physiological development \cite{brodschneider} and is believed to have considerable impact on longevity \cite{longevity}. At the colony level, pollen enables young workers to produce jelly, which then nourishes larvae, drones, older workers, and the queen \cite{jelly, jelly2}. Various environmental changes (including urbanization and monoculture crop production) have significantly altered the nutritional profile available to honey bees. In particular, honey bees are confronted with a less diverse selection of pollen, which is of concern because mixed-pollen (polyfloral) diets are generally considered healthier than single-pollen (monofloral) diets \cite{diverse, diverse2, alaux}. Indeed, reported colony mortality rates are higher in developed land areas compared to undeveloped land areas \cite{undeveloped}, and beekeepers rank poor nutrition as one of the main reasons for colony losses \cite{bkLoss}. Understanding how undiversified diets affect honey bee health will be crucial to resolve problems that may arise as agriculture continues to intensify throughout the world \cite{ag, ag2}.

Viral infection was a comparatively minor problem in honey bees until the last century when \textit{Varroa destructor} (an ectoparasitic mite) spread worldwide \cite{miteSpread}. This mite feeds on honey bee hemolymph \cite{hemolymph}, transmits multiple viruses, and supports replication of some viruses \cite{miteVirus, miteVirus2, miteVirus3, newVR}. More than 20 honey bee viruses have been identified \cite{numVirus}. One of these viruses that has been linked to honey bee decline is Israeli acute paralysis virus (IAPV). A positive-sense RNA virus of the family Dicistroviridae \cite{fam}, IAPV infection causes shivering wings, decreased locomotion, muscle spams, paralysis, and high premature death percentages in caged infected adult honey bees \cite{symptoms}. IAPV has demonstrated higher infectious capacities than other honey bee viruses under certain conditions \cite{carrillo} and is more prevalent in colonies that do not survive the winter \cite{winter}. Its role in the rising phenomenon of ``Colony Collapse Disorder'' (in which the majority of worker bees disappear from a hive) remains unclear: It has been implicated in some studies \cite{iapvCCD, iapvCCD2} but not in other studies \cite{ccd1, fam, iapvCCD3}. Nonetheless, it is clear that IAPV reduces colony strength and survival.

Although there is growing interest in how viruses and diet quality affect the health and sustainability of honey bees, as well as a recognition that such factors might operate interactively, there are only a small number of experimental studies thus far directed toward elucidating the interactive effects of these two factors in honey bees \cite{intNV, intNV2, intNV3}. We recently used laboratory cages and nucleus hive experiments to investigate the health effects of these two factors, and our results show the importance of the combined effects of both diet quality and virus infection. Specifically, ingestion by honey bees of high quality pollen is able to mitigate virus-induced mortality to the level of diverse, polyfloral pollen \cite{adamInt}.

Following up on these phenotypic findings from our previous study, we now aim to understand the corresponding underlying mechanisms by which high quality diets protect bees from virus-induced mortality. For example, it is not known whether the protective effect of good diet is due to direct, specific effects on immune function (resistance), or if it is due to indirect effects of good nutrition on vigor (tolerance) \cite{resTol1}. Transcriptomics is one means to better understand the mechanistic underpinnings of dietary and viral effects on honey bee health. Transcriptomic analysis can help us identify 1) the genomic scale of transcriptomic response to diet and virus infection, 2) whether these factors interact in an additive or synergistic way on transcriptome function, and 3) the types of pathways affected by diet quality and viral infection. This information, heretofore lacking in the literature, can help us better understand how good nutrition may be able to serve as a ``buffer'' against other stressors \cite{AdamTothReview}.

As it stands, there are only a small number of published experiments examining gene expression patterns related to diet effects \cite{alaux2} and virus infection effects \cite{galbraith} in honey bees. Honey bee transcriptomic studies have found that pollen nutrition upregulates genes involved in macromolecule metabolism, longetivity, and the insulin/TOR pathway required for physiological development \cite{alaux2}. Insect gene expression studies have implicated RNA silencing, autophagy, JAK/STAT, Toll, and IMD as antiviral pathways for a range of viral infections, including dicistrovirus infections \cite{galbraith, Avadhanula, fewDEGs2, kemp, costa}. Transcriptional pausing is also believed to be imperative for early antiviral immunity in many insects \cite{resTol11}. See \cite{cherryReview} for a review of known antiviral mechanisms in insect models.

As far as we know, there are few to no studies investigating honey bee gene expression patterns specifically related to monofloral diets, and few to no studies investigating honey bee gene expression patterns related to the combined effects of diet in any broad sense and viral inoculation in any broad sense. In this study, we examine how monofloral diets and viral inoculation influence gene expression patterns in honey bees by focusing on four treatment groups (low quality diet without IAPV exposure, high quality diet without IAPV exposure, low quality diet with IAPV exposure, and high quality diet with IAPV exposure). We conduct RNA-sequencing analysis on a randomly selected subset of the honey bees we used in our previous study (as is further described in our methods section). We then examine pairwise combinations of treatment groups, the main effect of monofloral diet, the main effect of IAPV exposure, and the combined effect of the two factors on gene expression patterns.

@@

As shown in Figure 2, IAPV titers of honey bees 72 hour post-inoculation significantly differed among the treatment groups (mixed model ANOVA across all treatment groups, df = 3, 33; F = 6.10; p $<$ 2.03e-03). The effect of virus treatment (mixed model ANOVA, df = 1, 33; F = 15.04; p $<$ 4.75e-04) was significant, but the diet treatment (mixed model ANOVA, df = 1, 33; F = 2.55; p = 1.20e-01) and the interaction between the two factors (mixed model ANOVA, df = 1, 33; F = 7.02e-01, p = 4.08e-01) were not significant. We compared IAPV titers  based on pairwise comparisons: Bees fed Rockrose pollen had significantly elevated IAPV titers with virus infection compared to uninfected controls (Benjamini Hochberg, p $<$ 7.56e-03). However, bees fed Chestnut pollen did not have significantly elevated IAPV titers with virus infection compared to uninfected controls (Benjamini Hochberg, p = 6.29e-02). Overall, we interpreted these findings to mean that high-quality Chestnut pollen could ``rescue'' high virus titers resulting from the inoculation treatment, whereas low-quality Rockrose pollen could not (Figure 2).

@@

We observed a substantially larger number of differentially expressed genes (DEGs) in our diet main effect (\textit{n} = 1,914) than in our virus main effect (\textit{n} = 43) (Supplementary table 1 A and B, Additional file 1). In the diet factor, more DEGs were upregulated in the more-nutritious Chestnut group (\textit{n} = 1,033) than in the less-nutritious Rockrose group (\textit{n} = 881). In the virus factor, there were more virus-upregulated DEGs (\textit{n} = 38) than control-upregulated DEGs (\textit{n} = 5). While these reported DEG counts are from the DESeq2 package, we saw similar trends for the edgeR and limma package results (Supplementary table 1, Additional file 1 and Additional file 18).

@@

We wished to explore the signal:to:noise ratio between the Galbraith dataset and our dataset. Note that the Galbraith dataset contained three samples for each virus level, while our dataset contained twelve samples for each virus level. Basic PCA plots were constructed with the DESeq2 analysis pipeline and showed that the Galbraith dataset may separate the infected and uninfected honey bees better than our dataset (Additional file 2). We also noted that the first replicate of both treatment groups in the Galbraith data did not cluster as cleanly in the PCA plots. However, through this automatically-generated plot, we can only visualize information at the sample level. Wanting to learn more about the data at the gene level, we continued with additional visualization techniques.

We used parallel coordinate lines superimposed onto boxplots to visualize the DEGs associated with virus infection in the two studies. The background side-by-side boxplot represents the distribution of all genes in the data, and each parallel coordinate line represents one DEG. To reduce overplotting of parallel coordinate lines, we used hierarchical clustering techniques to separate DEGs into common patterns as is described in the methods section.

@@

Finally, we looked at scatterplot matrices to assess the DEGs. We created standardized scatterplot matrices for each of the four clusters (from Figure 3) of the Galbraith data (Additional files 7, 8, 9, and 10). We also created standardized scatterplot matrices for our data. However, as our dataset contained 24 samples, we would need to include 276 scatterplots in our matrix, which would be too numerous to allow for efficient visual assessment of the data. As a result, we created four scatterplot matrices of our data, each with subsets of 6 samples to be more comparable to the Galbraith data (Additional files 11, 12, 13, and 14). We can again confirm through these plots that the DEGs from the Galbraith data appeared more as expected: They deviated more from the \textit{x=y} line in the treatment scatterplots while staying close to the \textit{x=y} line in replicate scatterplots.

@@

Using the contrasts specified in Table \ref{tbl:contrasts}, we discovered 122 ``tolerance'' candidate DEGs and 125 ``resistance'' candidate DEGs. We again used parallel coordinate lines superimposed onto side-by-side boxplots to visualize these candidate DEGs. To reduce overplotting of parallel coordinate lines, we again used hierarchical clustering techniques to separate DEGs into common patterns. Perhaps unsurprisingly, we still see a substantial amount of noise (inconsistency between replicates) in our resulting candidate DEGs (Additional files 15 and 16). However, the broad patterns we expect to see still emerge: For example, based on the contrasts we created to obtain the `tolerance'' candidate DEGs, we expect them to display larger count values in the ``NC'' group compared to the ``NR'' group and larger count values in the ``VC'' group compared to the ``VR'' group. Indeed, we see this pattern in the associated parallel coordinate plots (Additional file 15). Likewise, based on the contrasts we created to obtain the `resistance'' candidate DEGs, we still expect them to display larger count values in the ``VC'' group compared to the ``VR'' group, but we no longer expect to see a difference between the ``NC'' and ``NR'' groups. We do generally see these expected patterns in the associated parallel coordinate plots: While there are large outliers in the ``NC'' group, the ``NR'' replicates are no longer typically below a standardized count of zero (Additional file 16). The genes in Cluster 3 may follow the expected pattern the most distinctively (Additional file 16).

@@

In general, the R-squared values between gene read counts and pathogen response measurements were low (R-squared $<$ 0.1). However, some DEG clusters showed slightly larger R-squared values than the non-DEG group (the rest of the data). One prominent example of this includes the first and second cluster of the virus-related DEGs and their correlation with IAPV titers (Additional file 19I). The Kruskal–Wallis test was used to determine if R-squared populations of DEG clusters significantly differed from those in the rest of the data. The p-values and Bonferroni correction values for each of the 36 tests (as described in the methods section) is provided in Supplementary table 9, Additional file 1. An overall trend emerges to suggest that DEGs may have significantly larger correlation with the pathogen response measurements compared to non-DEGs. It is difficult to interpret these results in light of the noisiness of this data, but it may be of interest to conduct further studies examining differential expression between pathogen response measurements.

@@

While some insect systems have shown relatively low transcriptional responses to dicistrovirus infection \cite{fewDEGs, fewDEGs2}, previous work on honey bees has revealed many hundreds of DEGs \cite{galbraith}. Despite this, the transcriptomic response to virus infection in our experiment was fairly limited. We found only 43 transcripts to be differentially expressed, some with known immune functions (Table \ref{tbl:virusGenes}) such as argonaute-2 and a gene with similarity to MD-2 lipid recognition protein, as well as genes related to transcriptional regulation and muscle contraction. The small number of DEGs in this study may be partly explained by the large amount of noise in the data (Figure 4 and Additional files 2B, 4, 11, 12, 13, and 14).

@@

One of the goals of this study was to use our RNA-sequencing data to assess whether transcriptomic responses to diet quality and virus infection provide insight into whether high quality diet can buffer bees from pathogen stress via mechanisms of ``resistance'' or ``tolerance''. Recent evidence has suggested that overall immunity is determined by more than just ``resistance'' (the reduction of pathogen fitness within the host by mechanisms of avoidance and control) \cite{resTol2}. Instead, overall immunity is related to ``resistance'' in conjunction with ``tolerance'' (the reduction of adverse effects and disease resulting from pathogens by mechanisms of healing) \cite{resTol1, resTol2}. Immune-mediated resistance and diet-driven tolerance mechanisms are costly and may compete with each other \cite{resTol1, resTol4}. Data and models have suggested that selection can favor an optimum combination of both resistance and tolerance \cite{resTol5, resTol6, resTol7, resTol8}. We attempted to address this topic through specific gene expression contrasts (Table \ref{tbl:contrasts}), accompanied by GO analysis of the associated gene lists. We found an approximately equal number of resistance (\textit{n} = 125) and tolerance (\textit{n} = 122) related candidate DEGs, suggesting both processes may be playing significant roles in dietary buffering from pathogen induced mortality. Resistance candidate DEGs had functions related to several forms of metabolism (chitin and carbohydrate), regulation of transcription, and cell adhesion (Figure 5B). Tolerance candidate DEGs had functions related to carbohydrate metabolism and chitin metabolism; however, they also showed functions related to immune response, including RNA polymerase II transcription (Figure 5A). Previous studies have shown that transcriptional pausing of RNA polymerase II may be an innate immune response in \textit{D. melanogaster} that allows for a more rapid response by increasing the accessibility of promoter regions of virally induced genes \cite{resTol11}. These possible immunological defense mechanisms within our ``tolerance'' candidate DEGs and metabolic processes within our ``resistance'' candidate DEGs may provide additional evidence of feedbacks between diet and disease in honey bees \cite{AdamTothReview}.

@@

In addition, this study used a whole body RNA-sequencing approach. In future related studies, it may be informative to use tissue-specific methods. Previous work has shown that even though IAPV replication occurs in all honey bee tissues, it localizes more in gut and nerve tissues and in the hypopharyngeal glands. Likewise, the highest IAPV titers have been observed in gut tissues \cite{winter}. Recent evidence has suggested that RNA-sequencing approaches toward composite structures in honey bees leads to false negatives, implying that genes strongly differentially expressed in particular structures may not reach significance within the composite structure \cite{tissueLevel}. These studies have also found that within a composite extraction, structures therein may contain opposite patterns of differential expression. We can provide more detailed answers to our original transcriptomic questions if we were to repeat this same experimental design only now at a more refined tissue level. Another future direction related to this work would be to integrate multiple omics datasets to investigate monofloral diet quality and IAPV infection in honey bees. Indeed, previous studies in honey bees have found that multiple omics datasets do not always align in a clear-cut manner, and hence may broaden our understanding of the molecular mechanisms being explored \cite{galbraith}.

@@

Details of the procedures we used to prepare virus inoculum, infect and feed caged honey bees, and quantify IAPV can be reviewed in our previous work \cite{adamInt, carrillo}. A linear mixed effects model was used to relate the mortality rates and IAPV titers to the main and interaction effects of the diet and virus factors. The model was fitted to the data by restricted maximum likelihood (REML) using the ``lme'' function in the R package ``nlme''. A random (intercept) effect for experimental setup was included in the model. Post-hoc pairwise comparisons of the four (diet and virus combination) treatment groups were performed and Benjamini-Hochberg adjusted p-values were calculated to limit familywise Type I error rates \cite{bh}.

@@

We used an array of visualization tools as part of our analysis. We first used well-known tools like the PCA plot \cite{pca} from the DESeq2 package. After that, we used lesser-known multivariate visualization tools from our work-in-progress R package called bigPint. Specifically, we used parallel coordinate plots \cite{origPCP}, scatterplot matrices \cite{scatMat}, and litre plots (which we recently developed based on ``replicate line plots'' \cite{jds} (cite bigPint too)) to assess the variability between the replicates and the treatments in our data. We also used these plotting techniques to assess for normalization problems and other common problems in RNA-sequencing analysis pipelines \cite{jds} (cite bigPint too).
